---
title: "1_Follow_up"
output: html_document
date: "2025-04-22"
---

In this code I will directly address the points for follow-up that were denoted in the previous progress report.

The points are:
1. Confirm the genes of interest using the selection and filtering methods discussed.
2. Compare the LUSC genes of interest with the LUAD genes of interest.
3. Compare former vs. never smoker tumor methylation as another point of interest for the analysis.
4. Explore the statistical significance of overlapping genes and give a detailed report of the results and their implications.
5. Once we have confirmed that the lists of genes of interest are finalized, I can use the datasets listed in comment 8 to make further comparisons and refine the hypothesis.

Addressing the points:

# 1.  Confirm the genes of interest using the selection and filtering methods discussed.
This is up to the committee.

# 2. Compare the LUSC genes of interest with the LUAD genes of interest.

## Read in the genes of interest
```{r}

# LUAD 
A1_TE_merged_DEGs_75th_percentile_LUAD_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_merged_DEGs_75th_percentile_LUAD_CSFS_20250422.txt")

A1_TE_A2_linked_genes_75th_percentile_LUAD_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_A2_linked_genes_75th_percentile_LUAD_CSFS_20250414.txt")

# LUSC
A1_TE_merged_DEGs_75th_percentile_LUSC_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_merged_DEGs_75th_percentile_LUSC_CSFS_20250422.txt")

A1_TE_A2_linked_genes_75th_percentile_LUSC_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_A2_linked_genes_75th_percentile_LUSC_CSFS_20250414.txt")
```


## Compare the gene lists to each other with venn diagrams
```{r}

```


