---
title: "1_Follow_up"
output: html_document
date: "2025-04-22"
---

In this code I will directly address the points for follow-up that were denoted in the previous progress report.

The points are:
1. Confirm the genes of interest using the selection and filtering methods discussed.
2. Compare the LUSC genes of interest with the LUAD genes of interest.
3. Compare former vs. never smoker tumor methylation as another point of interest for the analysis.
4. Explore the statistical significance of overlapping genes and give a detailed report of the results and their implications.
5. Once we have confirmed that the lists of genes of interest are finalized, I can use the datasets listed in comment 8 to make further comparisons and refine the hypothesis.

Addressing the points:

# 1.  Confirm the genes of interest using the selection and filtering methods discussed.
This is up to the committee. When I change the lists of genes of interest it will affect the checks after this point.

# 2. Compare the LUSC genes of interest with the LUAD genes of interest.

## Read in the genes of interest
```{r}
## These are the genes of interest I have defined as of 2025/04/22

# LUAD 
A1_TE_merged_DEGs_75th_percentile_LUAD_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_merged_DEGs_75th_percentile_LUAD_CSFS_20250422.txt")

A1_TE_A2_linked_genes_75th_percentile_LUAD_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_A2_linked_genes_75th_percentile_LUAD_CSFS_20250414.txt")


# LUSC
A1_TE_merged_DEGs_75th_percentile_LUSC_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_merged_DEGs_75th_percentile_LUSC_CSFS_20250422.txt")

A1_TE_A2_linked_genes_75th_percentile_LUSC_CSFS <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_A2_linked_genes_75th_percentile_LUSC_CSFS_20250414.txt")

```


## Compare the gene lists to each other with euler/"venn" diagrams
```{r}
library(eulerr)

# A1_TE gene comparison between LUAD and LUSC

A1_TE_gene_sets <- euler(list(A1_TE_LUAD = unique(A1_TE_merged_DEGs_75th_percentile_LUAD_CSFS$Gene), A1_TE_LUSC = unique(A1_TE_merged_DEGs_75th_percentile_LUSC_CSFS$Gene)))

plot(main = "A1_TE genes",
     A1_TE_gene_sets,
     quantities = TRUE,  # Show counts in regions
     fills = c("red", "blue", "green"),  
     alpha = 0.5, 
     edges = TRUE)

# A1_TE_A2 comparison between LUAD and LUSC

A1_TE_A2_gene_sets <- euler(list(A1_TE_A2_LUAD = unique(A1_TE_A2_linked_genes_75th_percentile_LUAD_CSFS$Gene), A1_TE_A2_LUSC = unique(A1_TE_A2_linked_genes_75th_percentile_LUSC_CSFS$Gene)))

plot(main = "A1_TE_A2 genes",
     A1_TE_A2_gene_sets,
     quantities = TRUE,  # Show counts in regions
     fills = c("red", "blue", "green"),  
     alpha = 0.5, 
     edges = TRUE)

```


## Take a look at pathways for the genes that are common or different here
```{r}



```



# 3. Compare former vs. never smoker tumor methylation as another point of interest for the analysis.

I could see doing this in two different ways:

3.1. Take the A1-TE-A2 genes and check how well they can stratify former from never smokers in the TCGA methylation data using either a PCA or hierarchical clustering

3.2. Do differential analysis of former vs. never smokers in tumors and compare the gene list to the A1-TE-A2 list.

The first way sounds kinda more interesting and probably better, and I have already done something like it. I will try both of these anyway.

## 3.1. Take the A1-TE-A2 genes and check how well they can stratify current from former smokers in the TCGA methylation data using either a PCA or hierarchical clustering

### LUAD


#### Read in the clinical data 
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)

## Read in TCGA-LUSC clinical data
query_clinical <- GDCquery(
  project = "TCGA-LUAD",
  data.category = "Clinical",
  data.type = "Clinical Supplement"
)

# Check the available files
query_clinical$results[[1]] %>% head()

## I see that some are not BCR XML files, so I will try to remove these
query_clinical$results[[1]] <- query_clinical$results[[1]] %>%
  filter(data_format == "BCR XML")

## Download data
GDCdownload(query_clinical,files.per.chunk = 1)
clinical_data <- GDCprepare_clinic(query_clinical, clinical.info = "patient")

## Sort by tobacco smoking history
clinical_data_smoking_history <- clinical_data %>%
  arrange(., tobacco_smoking_history) %>%
  select(bcr_patient_barcode, tobacco_smoking_history)

```



### Read in the methylation data and arrange by smoking history
```{r}

# Get the patient barcodes in a form that can be used to filter the tumor samples
clinical_data_smoking_history_sample_barcodes <- clinical_data_smoking_history %>%
  mutate(bcr_patient_barcode = bcr_patient_barcode %>%
    gsub("$", ".01", .) %>%
    gsub("-", "\\.", .)
  )

## Read in the data for TCGA-LUAD methylation
LUAD_methylation_hm450_tumor <- read.table("../../4_TCGA_data/TCGA_LUAD/data_methylation_hm450_LUAD.txt", header=TRUE, fill=TRUE)

# Get the barcodes existing in rownames of the sample info
clinical_data_smoking_history_sample_barcodes_methyl <- clinical_data_smoking_history_sample_barcodes[clinical_data_smoking_history_sample_barcodes$bcr_patient_barcode %in% colnames(LUAD_methylation_hm450_tumor),] 

```


## Deal with duplicate genes and make them the rownames of the methylation table
```{r}
### 3.2.2 Handling duplicate genes
library(dplyr)

# Step 1: Identify rows with duplicate Gene values
duplicates <- LUAD_methylation_hm450_tumor %>%
  group_by(Hugo_Symbol) %>%
  filter(n() > 1)

# Step 2: Remove the row with the smallest row sum for each duplicate gene
rows_to_remove <- duplicates %>%
  rowwise() %>%
  mutate(row_sum = sum(c_across(where(is.numeric)))) %>%  # Compute row sum for numeric columns
  group_by(Hugo_Symbol) %>%
  slice_min(row_sum, with_ties = FALSE) %>%  # Select the row with the smallest sum
  ungroup()

# Step 3: Remove these rows from the original dataframe
LUAD_methylation_hm450_tumor <- anti_join(LUAD_methylation_hm450_tumor, rows_to_remove, by = colnames(LUAD_methylation_hm450_tumor))

# Now make the gene names into row names
rownames(LUAD_methylation_hm450_tumor) <- LUAD_methylation_hm450_tumor$Hugo_Symbol

# Remove the ID columns
LUAD_methylation_hm450_tumor$Hugo_Symbol <- NULL
LUAD_methylation_hm450_tumor$Entrez_Gene_Id <- NULL

```


## PCA of the LUAD methylation data using A1-TE genes and coloring by smoking status (filtering out 5 (FS of unspecified TSQ duration)) 
```{r}
library(limma)

## Setting up the dataframe

# Order the dataframe by smoking status
LUAD_methylation_hm450_tumor_sstat <- LUAD_methylation_hm450_tumor %>%
  select(clinical_data_smoking_history_sample_barcodes_methyl$bcr_patient_barcode)

# Filter to just A1-TE genes
LUAD_methylation_hm450_tumor_sstat_A1_TE <- LUAD_methylation_hm450_tumor_sstat %>%
  filter(rownames(.) %in% A1_TE_merged_DEGs_75th_percentile_LUAD_CSFS$Gene)


## Plot a PCA based on A1-TE genes, using color to represent smoking status

# Smoking status
smoking_status <- clinical_data_smoking_history_sample_barcodes_methyl$tobacco_smoking_history

# Only keep samples with status 1, 2, 3, or 4
valid_idx <- smoking_status %in% c(1, 2, 3, 4)
smoking_status_filtered <- smoking_status[valid_idx]

# Define custom order for color darkness
status_order <- c(1, 3, 4, 2)  # lightest to darkest

# Create color ramp
blues <- colorRampPalette(c("#cce6ff", "#004080"))(length(status_order))

# Create a named vector mapping statuses to colors
status_color_map <- setNames(blues, status_order)

# Assign colors to filtered samples
colors <- status_color_map[as.character(smoking_status_filtered)]

# Filter the data for plotting
filtered_data <- LUAD_methylation_hm450_tumor_sstat_A1_TE[, valid_idx]

# Plot MDS
plotMDS(filtered_data,
        gene.selection = "common",
        main = "PCA for TCGA-LUAD methylation by smoking status using A1-TE genes",
        col = colors,
        pch = 1)

# Add legend
legend("bottomleft",
       legend = paste(c("1 = NS",  "3 = FS > 15 yr", "4 = FS≤ 15 years", "2 = CS")),
       col = blues,
       pch = 1,
       title = "Smoking History")

## Doesn't look like much of a trend happening here

```


## PCA of the LUAD methylation data using A1-TE-A2 genes and coloring by smoking status (filtering out 5 (FS of unspecified TSQ duration)) 
```{r}
library(limma)

## Setting up the dataframe

# Order the dataframe by smoking status
LUAD_methylation_hm450_tumor_sstat <- LUAD_methylation_hm450_tumor %>%
  select(clinical_data_smoking_history_sample_barcodes_methyl$bcr_patient_barcode)

# Filter to just A1-TE-A2 genes
LUAD_methylation_hm450_tumor_sstat_A1_TE_A2 <- LUAD_methylation_hm450_tumor_sstat %>%
  filter(rownames(.) %in% A1_TE_A2_linked_genes_75th_percentile_LUAD_CSFS$Gene)



## Plot a PCA based on A1-TE-A2 genes, using color to represent smoking status

# Smoking status
smoking_status <- clinical_data_smoking_history_sample_barcodes_methyl$tobacco_smoking_history

# Only keep samples with status 1, 2, 3, or 4
valid_idx <- smoking_status %in% c(1, 2, 3, 4)
smoking_status_filtered <- smoking_status[valid_idx]

# Define custom order for color darkness
status_order <- c(1, 3, 4, 2)  # lightest to darkest

# Create color ramp
blues <- colorRampPalette(c("#cce6ff", "#004080"))(length(status_order))

# Create a named vector mapping statuses to colors
status_color_map <- setNames(blues, status_order)

# Assign colors to filtered samples
colors <- status_color_map[as.character(smoking_status_filtered)]

# Filter the data for plotting
filtered_data <- LUAD_methylation_hm450_tumor_sstat_A1_TE_A2[, valid_idx]

# Plot MDS
plotMDS(filtered_data,
        gene.selection = "common",
        main = "PCA for TCGA-LUAD methylation by smoking status using A1-TE-A2 genes",
        col = colors,
        pch = 1)

# Add legend
legend("topright",
       legend = paste(c("1 = NS",  "3 = FS > 15 yr", "4 = FS≤ 15 years", "2 = CS")),
       col = blues,
       pch = 1,
       title = "Smoking History")

## Doesn't look like much of a trend happening here

```


#########

### LUSC

#### Read in the clinical data 
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
library(dplyr)

## Read in TCGA-LUSC clinical data
query_clinical <- GDCquery(
  project = "TCGA-LUSC",
  data.category = "Clinical",
  data.type = "Clinical Supplement"
)

# Check the available files
query_clinical$results[[1]] %>% head()

## I see that some are not BCR XML files, so I will try to remove these
query_clinical$results[[1]] <- query_clinical$results[[1]] %>%
  filter(data_format == "BCR XML")

## Download data
GDCdownload(query_clinical,files.per.chunk = 1)
clinical_data <- GDCprepare_clinic(query_clinical, clinical.info = "patient")

## Sort by tobacco smoking history
clinical_data_smoking_history <- clinical_data %>%
  arrange(., tobacco_smoking_history) %>%
  select(bcr_patient_barcode, tobacco_smoking_history)

```


### Read in the methylation data and arrange by smoking history
```{r}

# Get the patient barcodes in a form that can be used to filter the tumor samples
clinical_data_smoking_history_sample_barcodes <- clinical_data_smoking_history %>%
  mutate(bcr_patient_barcode = bcr_patient_barcode %>%
    gsub("$", ".01", .) %>%
    gsub("-", "\\.", .)
  )

## Read in the data for TCGA-LUSC methylation
LUSC_methylation_hm450_tumor <- read.table("../../4_TCGA_data/TCGA_LUSC/data_methylation_hm450_LUSC.txt", header=TRUE, fill=TRUE)

# Get the barcodes existing in rownames of the sample info
clinical_data_smoking_history_sample_barcodes_methyl <- clinical_data_smoking_history_sample_barcodes[clinical_data_smoking_history_sample_barcodes$bcr_patient_barcode %in% colnames(LUSC_methylation_hm450_tumor),] 

```

## Deal with duplicate genes and make them the rownames of the methylation table
```{r}
### 3.2.2 Handling duplicate genes
library(dplyr)

# Step 1: Identify rows with duplicate Gene values
duplicates <- LUSC_methylation_hm450_tumor %>%
  group_by(Hugo_Symbol) %>%
  filter(n() > 1)

# Step 2: Remove the row with the smallest row sum for each duplicate gene
rows_to_remove <- duplicates %>%
  rowwise() %>%
  mutate(row_sum = sum(c_across(where(is.numeric)))) %>%  # Compute row sum for numeric columns
  group_by(Hugo_Symbol) %>%
  slice_min(row_sum, with_ties = FALSE) %>%  # Select the row with the smallest sum
  ungroup()

# Step 3: Remove these rows from the original dataframe
LUSC_methylation_hm450_tumor <- anti_join(LUSC_methylation_hm450_tumor, rows_to_remove, by = colnames(LUSC_methylation_hm450_tumor))

# Now make the gene names into row names
rownames(LUSC_methylation_hm450_tumor) <- LUSC_methylation_hm450_tumor$Hugo_Symbol

# Remove the ID columns
LUSC_methylation_hm450_tumor$Hugo_Symbol <- NULL
LUSC_methylation_hm450_tumor$Entrez_Gene_Id <- NULL

```



## PCA of the LUSC methylation data using A1-TE genes and coloring by smoking status (filtering out 5 (FS of unspecified TSQ duration)) 
```{r}
library(limma)

## Setting up the dataframe

# Order the dataframe by smoking status
LUSC_methylation_hm450_tumor_sstat <- LUSC_methylation_hm450_tumor %>%
  select(clinical_data_smoking_history_sample_barcodes_methyl$bcr_patient_barcode)

# Filter to just A1-TE genes
LUSC_methylation_hm450_tumor_sstat_A1_TE <- LUSC_methylation_hm450_tumor_sstat %>%
  filter(rownames(.) %in% A1_TE_merged_DEGs_75th_percentile_LUSC_CSFS$Gene)


## Plot a PCA based on A1-TE genes, using color to represent smoking status

# Smoking status
smoking_status <- clinical_data_smoking_history_sample_barcodes_methyl$tobacco_smoking_history

# Only keep samples with status 1, 2, 3, or 4
valid_idx <- smoking_status %in% c(1, 2, 3, 4)
smoking_status_filtered <- smoking_status[valid_idx]

# Define custom order for color darkness
status_order <- c(1, 3, 4, 2)  # lightest to darkest

# Create color ramp
blues <- colorRampPalette(c("#cce6ff", "#004080"))(length(status_order))

# Create a named vector mapping statuses to colors
status_color_map <- setNames(blues, status_order)

# Assign colors to filtered samples
colors <- status_color_map[as.character(smoking_status_filtered)]

# Filter the data for plotting
filtered_data <- LUSC_methylation_hm450_tumor_sstat_A1_TE[, valid_idx]

# Plot MDS
plotMDS(filtered_data,
        gene.selection = "common",
        main = "PCA for TCGA-LUSC methylation by smoking status using A1-TE genes",
        col = colors,
        pch = 1)

# Add legend
legend("bottomleft",
       legend = paste(c("1 = NS",  "3 = FS > 15 yr", "4 = FS≤ 15 years", "2 = CS")),
       col = blues,
       pch = 1,
       title = "Smoking History")

## Doesn't look like much of a trend happening here

```


## PCA of the LUSC methylation data using A1-TE-A2 genes and coloring by smoking status (filtering out 5 (FS of unspecified TSQ duration)) 
```{r}
library(limma)

## Setting up the dataframe

# Order the dataframe by smoking status
LUSC_methylation_hm450_tumor_sstat <- LUSC_methylation_hm450_tumor %>%
  select(clinical_data_smoking_history_sample_barcodes_methyl$bcr_patient_barcode)

# Filter to just A1-TE-A2 genes
LUSC_methylation_hm450_tumor_sstat_A1_TE_A2 <- LUSC_methylation_hm450_tumor_sstat %>%
  filter(rownames(.) %in% A1_TE_A2_linked_genes_75th_percentile_LUSC_CSFS$Gene)



## Plot a PCA based on A1-TE-A2 genes, using color to represent smoking status

# Smoking status
smoking_status <- clinical_data_smoking_history_sample_barcodes_methyl$tobacco_smoking_history

# Only keep samples with status 1, 2, 3, or 4
valid_idx <- smoking_status %in% c(1, 2, 3, 4)
smoking_status_filtered <- smoking_status[valid_idx]

# Define custom order for color darkness
status_order <- c(1, 3, 4, 2)  # lightest to darkest

# Create color ramp
blues <- colorRampPalette(c("#cce6ff", "#004080"))(length(status_order))

# Create a named vector mapping statuses to colors
status_color_map <- setNames(blues, status_order)

# Assign colors to filtered samples
colors <- status_color_map[as.character(smoking_status_filtered)]

# Filter the data for plotting
filtered_data <- LUSC_methylation_hm450_tumor_sstat_A1_TE_A2[, valid_idx]

# Plot MDS
plotMDS(filtered_data,
        gene.selection = "common",
        main = "PCA for TCGA-LUSC methylation by smoking status using A1-TE-A2 genes",
        col = colors,
        pch = 1)

# Add legend
legend("topright",
       legend = paste(c("1 = NS",  "3 = FS > 15 yr", "4 = FS≤ 15 years", "2 = CS")),
       col = blues,
       pch = 1,
       title = "Smoking History")

## Doesn't look like much of a trend happening here

```


So far it doesn't look like there is much of a trend happening here. I could also try hierarchical clustering, and maybe also try straight differential analysis and gene comparison instead. Also, maybe the thing that Carl DeBoer suggested which is looking at the "loads" of the genes in the PCA?


##4. Explore the statistical significance of overlapping genes and give a detailed report of the results and their implications.

For my foray into this see the script hypergeometric_confusion and the word document "SOS question".



