---
title: "Hypergeometric_confusion"
output: html_document
date: "2025-04-01"
---

I am confused about my application of the hypergeometric test to check different filtering steps. I will explain here:

## Reading in datasets to intersect
```{r}

# A1 (airway expression in brushing samples comparing current and never smokers)
GSE63127_CS_NS   <- read.table("../2_Outputs/1_Airway_DEGs/GSE63127_CS_NS_limma_20250307.txt", header = TRUE)

# T-E (tumor expression comparing tumor and normal tissue (lung adenocarcinoma from TCGA))
TCGA_LUAD_DEG    <- read.table("../2_Outputs/4_Tumor_DEGs/TCGA_LUAD_limma_DEG_CSFS_20250331.txt")

```

## Using percentiles for filtering
```{r, results = 'hold'}
library(dplyr)

print("First format them all the same way (this time not filtering for FDR)")
GSE63127_CS_NS_formatted <- GSE63127_CS_NS %>%
  dplyr::rename(., Gene = Gene.symbol, FDR_A1 = adj.P.Val, log2FC_A1 = logFC) %>%
  dplyr::select(., Gene, FDR_A1, log2FC_A1)

TCGA_LUAD_DEG_formatted<- TCGA_LUAD_DEG %>%
  dplyr::rename(., FDR_TE = FDR, log2FC_TE = log2FC, Gene = Gene) %>%
  dplyr::select(., Gene, FDR_TE, log2FC_TE)

#####

print("Then filter to shared gene symbols")
GSE63127_CS_NS_A1_TE_shared_genes <- GSE63127_CS_NS_formatted %>%
  dplyr::filter(Gene %in% TCGA_LUAD_DEG$Gene)
print("GSE63127_CS_NS_A1_TE_shared_genes:")
nrow(GSE63127_CS_NS_A1_TE_shared_genes)

TCGA_LUAD_DEG_A1_TE_shared_genes <- TCGA_LUAD_DEG_formatted %>%
  dplyr::filter(Gene %in% GSE63127_CS_NS$Gene)
print("TCGA_LUAD_DEG_A1_TE_shared_genes:")
nrow(TCGA_LUAD_DEG_A1_TE_shared_genes)

#####

print("Then filter to FDRs individually being <= 0.05")
GSE63127_CS_NS_A1_TE_sig <- GSE63127_CS_NS_A1_TE_shared_genes %>%
  filter(FDR_A1 <= 0.05)
print("GSE63127_CS_NS_A1_TE_sig:")
nrow(GSE63127_CS_NS_A1_TE_sig)

TCGA_LUAD_DEG_A1_TE_sig <- TCGA_LUAD_DEG_A1_TE_shared_genes %>%
  filter(FDR_TE <= 0.05)
print("TCGA_LUAD_DEG_A1_TE_sig:")
nrow(TCGA_LUAD_DEG_A1_TE_sig)

#####

print("Then filter to FDR < 0.05 in both")
GSE63127_A1_TE_sig_all <- GSE63127_CS_NS_A1_TE_sig %>%
  filter(Gene %in% TCGA_LUAD_DEG_A1_TE_sig$Gene)
print("GSE63127_A1_TE_sig_all:")
nrow(GSE63127_A1_TE_sig_all)

TCGA_LUAD_DEG_A1_TE_sig_all <- TCGA_LUAD_DEG_A1_TE_sig %>%
  filter(Gene %in% GSE63127_CS_NS_A1_TE_sig$Gene)
print("TCGA_LUAD_DEG_A1_TE_sig_all:")
nrow(TCGA_LUAD_DEG_A1_TE_sig_all)

#####

print("Then filter to n% most up/downregulated genes based on log2FC.")
percentile <- 4
print(paste0(percentile, " percentile"))

##A1:
# Calculate percentile cutoffs
top_cutoff_A1 <- quantile(GSE63127_A1_TE_sig_all$log2FC_A1, 0.5+percentile/100/2, na.rm = TRUE)
bottom_cutoff_A1 <- quantile(GSE63127_A1_TE_sig_all$log2FC_A1, 0.5-percentile/100/2, na.rm = TRUE)

# Filter genes in the top and bottom n% of log2FC
GSE63127_CS_NS_percentile <- GSE63127_A1_TE_sig_all %>%
  filter(log2FC_A1 >= top_cutoff_A1 | log2FC_A1 <= bottom_cutoff_A1)
print("GSE63127_CS_NS_percentile:")
nrow(GSE63127_CS_NS_percentile)

## TE
# Calculate percentile cutoffs
top_cutoff_TE <- quantile(TCGA_LUAD_DEG_A1_TE_sig_all$log2FC_TE,  0.5+percentile/100/2, na.rm = TRUE)  # Top n%
bottom_cutoff_TE <- quantile(TCGA_LUAD_DEG_A1_TE_sig_all$log2FC_TE, 0.5-percentile/100/2, na.rm = TRUE)  # Bottom n%

# Filter genes in the top and bottom n% of log2FC
TCGA_LUAD_DEG_percentile <- TCGA_LUAD_DEG_A1_TE_sig_all %>%
  filter(log2FC_TE >= top_cutoff_TE | log2FC_TE <= bottom_cutoff_TE)
print("TCGA_LUAD_DEG_percentile:")
nrow(TCGA_LUAD_DEG_percentile)

```

## First trying the hypergeometric test at the FDR filter point (FDR in each vs FDR in all)

```{r, results = 'hold'}
library(SuperExactTest)

# Define gene lists as those with FDR < 0.05 in each individual dataset
A1_genes <- GSE63127_CS_NS_A1_TE_sig$Gene
TE_genes <- TCGA_LUAD_DEG_A1_TE_sig$Gene

# Universe of genes (e.g., all gene symbols shared between the two datasets)
universe <- nrow(GSE63127_CS_NS_A1_TE_shared_genes) 

# Create a list of gene sets
gene_sets <- list(A1_genes, TE_genes)

# Perform SuperExactTest
res <- supertest(gene_sets, n = universe)

# Print results
summary(res)

# Summarize results in a more readable format
df_res <- as.data.frame(summary(res)$Table)

## At this stage it definitely looks like chance.

```

## Now try the hypergeometric test again, but here specifically testing the percentile filter

```{r, results = 'hold'}
library(SuperExactTest)

# Define gene lists
A1_genes <- GSE63127_CS_NS_percentile$Gene
TE_genes <- TCGA_LUAD_DEG_percentile$Gene

# Universe of genes (e.g., all genes in the study - 
# So if we are comparing genes overlapping after the percentile filtering step, we should be using the list from right before that: symbols shared between A1 and TE with FDR < 0.05 in both, therefore with chance of overlap at this next step)
universe <- nrow(GSE63127_A1_TE_sig_all) 

# Create a list of gene sets
gene_sets <- list(A1_genes, TE_genes)

# Perform SuperExactTest
res <- supertest(gene_sets, n = universe)

# Print results
summary(res)

# Summarize results in a more readable format
df_res <- as.data.frame(summary(res)$Table)

# If we test this step, it is indeed significant. Does testing this step violate any part of the assumptions of the hyupergeometric test? Unsure.

```


## Then do overlaps as we see fit

```{r}

## Merge A1 and TE
A1_TE_merged_DEGs <- GSE63127_CS_NS_percentile %>%
  inner_join(., TCGA_LUAD_DEG_percentile, by = "Gene") %>%
  filter(sign(log2FC_A1)==sign(log2FC_TE)) # Filter to genes with same signs
nrow(A1_TE_merged_DEGs)

```



Just for the heck of it I want to try using the other package that Lies showed to me instead of SuperExactTest.

```{r}
# Testing the shared FDR < 0.05 filter step
N = nrow(GSE63127_CS_NS_A1_TE_shared_genes)  # All genes with symbols shared
m <- length(GSE63127_CS_NS_A1_TE_sig$Gene)   # Airway genes with FDR < 0.05
n <- length(TCGA_LUAD_DEG_A1_TE_sig$Gene)   # Tumor genes with FDR < 0.05
k <- length(intersect(GSE63127_CS_NS_A1_TE_sig$Gene, TCGA_LUAD_DEG_A1_TE_sig$Gene))    # Observed overlap
p_value <- phyper(k - 1, m, N - m, n, lower.tail = FALSE)
p_value # About the same as what I got before, not at all significant


# Testing the percentile filter step
N <- nrow(GSE63127_A1_TE_sig_all)  # All genes with FDR < 0.05 in both airway and tumor
m <- length(GSE63127_CS_NS_percentile$Gene)   # Airway genes with FDR < 0.05 in both airway and tumor and top 25% |log2FC|
n <- length(TCGA_LUAD_DEG_percentile$Gene)   # Tumor genes with FDR < 0.05 in both airway and tumor and top 25% |log2FC|
k <- length(intersect(GSE63127_CS_NS_percentile$Gene, TCGA_LUAD_DEG_percentile$Gene))    # Observed overlap
p_value <- phyper(k - 1, m, N - m, n, lower.tail = FALSE)
p_value # About the same as what I got before, extremely significant




```





