---
title: "suggestion_tries"
output: html_document
date: "2025-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Carl DeBoer suggested mashing datasets together and plotting a big PCA. I think I'll first try this with A1 and TE for LUAD.

# Multi-Dataset PCA (DeBoer's suggestion)

## Read in data
```{r}

## Read in the A1 normalized expression data
A1_exprs <- read.table("../../../Airway_expression/A1_exprs_20250403.txt", sep = " ")

## Read in the TE normalized expression data
TE_exprs <- read.table("../../../Tumor_expression/TE_data_20250408.txt", sep = " ")

```

## Handle duplicates 
```{r}
## Keep rows with highest sums

TE_exprs <- TE_exprs %>%
  mutate(row_sum = rowSums(across(where(is.numeric)))) %>%
  group_by(Gene) %>%
  slice_max(order_by = row_sum, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-row_sum)
TE_exprs <- column_to_rownames(TE_exprs, var = "Gene")

A1_exprs <- A1_exprs %>%
  mutate(row_sum = rowSums(across(where(is.numeric)))) %>%
  group_by(Gene) %>%
  slice_max(order_by = row_sum, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(-row_sum)
A1_exprs <- column_to_rownames(A1_exprs, var = "Gene")

```


# Trying Alejandro/Lies' suggestion of checking whether genes of interest are biased towards high read count
## Checking this bias with an MA plot for the TE dataset
```{r}
library(DESeq2)

groups <- rep(c("Tumor", "Normal"), length(TE_exprs)/2)

design <- model.matrix(~0 + groups)

colData <- data.frame(sample = colnames(TE_exprs), row.names = colnames(TE_exprs), group = groups)

dds <- DESeqDataSetFromMatrix(countData = as.matrix(TE_exprs),
                              colData = colData,
                              design= design)

ds <- DESeq(dds)
res <- results(ds)

# MA plot
plotMA(res, ylim = c(-5, 25))
```



# Trying DeBoer's suggestion of massive PCA
## Combine datasets together
```{r}
TE_exprs_filtered <- TE_exprs_nodup %>%
  filter(Gene %in% A1_exprs_nodup$Gene)

A1_exprs_filtered <- A1_exprs_nodup %>%
  filter(Gene %in% TE_exprs_filtered$Gene)

A1_TE_exprs <- inner_join(A1_exprs_filtered, TE_exprs_filtered, by = "Gene")

A1_TE_exprs <- column_to_rownames(A1_TE_exprs, var = "Gene")


## Hmm there are some negative values, check on that
A1_TE_exprs_negatives_only <- A1_TE_exprs %>%
  select(where(~ any(. < 0, na.rm = TRUE)))
## It's just for a few genes, I think I'll just remove them...
A1_TE_exprs <- A1_TE_exprs %>%
  filter(!rownames(.) %in% c("RTKN2", "LRRN4", "ADGRF4", "SLC22A17"))
```


## Big PCA and heatmap
```{r}
library(DESeq2)
library("pheatmap")
library("vsn")
library(gridExtra)
library(grid)

# Create a design matrix

groups <- rep(c("Current_smoker","Never_smoker"), length(A1_exprs)/2)
groups <- c(groups, rep(c("Tumor", "Normal"), length(TE_exprs)/2))

design <- model.matrix(~0 + groups)

colData <- data.frame(sample = colnames(A1_TE_exprs), row.names = colnames(A1_TE_exprs), group = groups)

dds <- DESeqDataSetFromMatrix(countData = as.matrix(A1_TE_exprs),
                              colData = colData,
                              design= design)


# Variance stabilizing transformations (VST) for visualization
vsd <- vst(dds, blind=FALSE)
#head(assay(vsd), 3)
plotPCA(vsd, intgroup=c("group"))

# Normalize and check variance
# this gives log2(n + 1)
ntd <- normTransform(dds)
meanSdPlot(assay(ntd))
meanSdPlot(assay(vsd))


# Heatmap of count matrix
dds <- estimateSizeFactors(dds)  # Estimates size factors for normalization
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:1000]

df <- as.data.frame(groups)
rownames(df) <- rownames(colData)

# Order columns based on group
ordered_cols <- order(colData$group)

# Heatmap
p <- pheatmap(assay(ntd)[select, ordered_cols], 
         cluster_rows=TRUE,
         cluster_cols=TRUE, 
         show_rownames=FALSE,
         show_colnames = TRUE,
         treeheight_row = 0,
         treeheight_col = 0,
         gaps_col = 1,
         annotation_col=df[ordered_cols, , drop=FALSE])

# TCGA-38-4626-01A still appears as an outlier in the heatmap even if that disappears in the PCA
# Note that the samples with suffix -01B cluster together and look quite different. These were FFPE. Potentially remove?

```

#####

Lies suggested checking whether the overlapping genes are biased towards the high read count levels.

# Checking read count bias of overlapping genes

## Read in the necessary gene lists and datasets
```{r}

# Read in the lists of DEGs for the A1 and TE individually
GSE63127_CS_NS_limma <- read.table("../../2_Outputs/1_Airway_DEGs/GSE63127_CS_NS_limma_20250307.txt")
TCGA_LUAD_limma_DEG_CSFS <- read.table("../../2_Outputs/2_Tumor_DEGs/TCGA_LUAD_limma_DEG_CSFS_20250331.txt")

# Read in the lists of overlapping genes
A1_TE_merged_DEGs <- read.table("../../2_Outputs/4_Linked_genes/A1_TE_merged_DEGs_25th_percentile_20250414.txt")
A1_TE_A2_linked_genes <- read.table( "../../2_Outputs/4_Linked_genes/A1_TE_A2_linked_genes_25th_percentile_20250414.txt")

# Read in the expression data
## Read in the A1 normalized expression data
A1_exprs <- read.table("../../../Airway_expression/A1_exprs_20250403.txt", sep = " ")

## Read in the TE normalized expression data
TE_exprs <- read.table("../../../Tumor_expression/TE_data_20250408.txt", sep = " ")

```

# Get mean expression values for the genes, add them to the DEG list
```{r}

A1_exprs_noblanks <- A1_exprs[A1_exprs$Gene != "",]
A1_exprs_means <-data.frame(Gene = A1_exprs_noblanks$Gene, mean_expr = rowMeans(as.matrix(A1_exprs_noblanks[,-1])))
A1_exprs_means$ID <- rownames(A1_exprs_means)

TE_exprs_noblanks <- TE_exprs[TE_exprs$Gene != "",]
TE_exprs_means <- data.frame(Gene = TE_exprs_noblanks$Gene, mean_expr = rowMeans(as.matrix(TE_exprs_noblanks[,-1])))


## Add the mean expression info to the DEGs lists
GSE63127_CS_NS_limma <- inner_join(GSE63127_CS_NS_limma, A1_exprs_means, by = "ID" )
TCGA_LUAD_limma_DEG_CSFS  <- inner_join(TCGA_LUAD_limma_DEG_CSFS, TE_exprs_means, by = "Gene")

```

# Add columns saying whether these are in the A1-TE overlap
```{r}
GSE63127_CS_NS_limma$A1_TE <- ifelse(GSE63127_CS_NS_limma$Gene %in% A1_TE_merged_DEGs$Gene, "Yes", "No")
GSE63127_CS_NS_limma$A1_TE_A2 <- ifelse(GSE63127_CS_NS_limma$Gene %in% A1_TE_A2_linked_genes$Gene, "Yes", "No")

TCGA_LUAD_limma_DEG_CSFS$A1_TE <- ifelse(TCGA_LUAD_limma_DEG_CSFS$Gene %in% A1_TE_merged_DEGs$Gene, "Yes", "No")
TCGA_LUAD_limma_DEG_CSFS$A1_TE_A2 <- ifelse(TCGA_LUAD_limma_DEG_CSFS$Gene %in% A1_TE_A2_linked_genes$Gene, "Yes", "No")
```


## Make an MA plot for TE
```{r}
library(ggplot2)

ggplot(TCGA_LUAD_limma_DEG_CSFS, aes(x = log10(mean_expr + 1), y = log2FC, color = A1_TE)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Yes" = "blue", "No" = "grey")) +
  theme_minimal() +
  labs(
    title = "MA Plot for TE, checking A1-TE genes",
    x = "Log10(Mean Expression + 1)",
    y = "Log2 Fold Change"
  ) +
  theme(legend.title = element_blank())

ggplot(TCGA_LUAD_limma_DEG_CSFS, aes(x = log10(mean_expr + 1), y = log2FC, color = A1_TE_A2)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Yes" = "blue", "No" = "grey")) +
  theme_minimal() +
  labs(
    title = "MA Plot for TE, checking A1-TE-A2 genes",
    x = "Log10(Mean Expression + 1)",
    y = "Log2 Fold Change"
  ) +
  theme(legend.title = element_blank())

```

It appears that there is not a bias towards high mean expression here. Good!


## Make an MA plot for A1
```{r}
library(ggplot2)

ggplot(GSE63127_CS_NS_limma, aes(x = log10(mean_expr + 1), y = logFC, color = A1_TE)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Yes" = "blue", "No" = "grey")) +
  theme_minimal() +
  labs(
    title = "MA Plot for A1, checking A1-TE genes",
    x = "Log10(Mean Expression + 1)",
    y = "Log2 Fold Change"
  ) +
  theme(legend.title = element_blank())

ggplot(GSE63127_CS_NS_limma, aes(x = log10(mean_expr + 1), y = logFC, color = A1_TE_A2)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c("Yes" = "blue", "No" = "grey")) +
  theme_minimal() +
  labs(
    title = "MA Plot for A1, checking A1-TE-A2 genes",
    x = "Log10(Mean Expression + 1)",
    y = "Log2 Fold Change"
  ) +
  theme(legend.title = element_blank())
```




