---
title: "Other dataset comparisons"
output: html_document
date: "2025-05-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in lists of genes of interest
```{r}
## LUAD
LUAD_A1_TE <- readLines("../../2_Outputs/4_Linked_genes/LUAD_A1_TE_genes_20250514.txt")
LUAD_A1_TE_A2 <- readLines("../../2_Outputs/4_Linked_genes/LUAD_A1_TE_A2_genes_20250514.txt")
LUAD_A1_TE_A2_TM <- readLines("../../2_Outputs/4_Linked_genes/LUAD_A1_TE_A2_TM_genes_20250528.txt")

## LUSC
LUSC_A1_TE <- readLines("../../2_Outputs/4_Linked_genes/LUSC_A1_TE_genes_20250514.txt")
LUSC_A1_TE_A2 <- readLines("../../2_Outputs/4_Linked_genes/LUSC_A1_TE_A2_genes_20250514.txt")
LUSC_A1_TE_A2_TM <- readLines("../../2_Outputs/4_Linked_genes/LUSC_A1_TE_A2_TM_genes_20250528.txt")

```


# Airway epithelial cell lines exposed to cigarette smoke

## Study 1: PMID: 18694499
This study looks at cell lines exposed to cigarette smoke and then which ones persist over different timepoints up to 24 hours.
```{r}
library(openxlsx)
library(dplyr)

sheet_names <- getSheetNames("../2_Gene_lists/cell_lines/18694499/18694499_S1.xlsx")
df_list <- lapply(sheet_names, function(s) read.xlsx("../2_Gene_lists/cell_lines/18694499/18694499_S1.xlsx", sheet = s))
names(df_list) <- sheet_names

# Right now I'm, particularly interested in the 24 hr ones since this is the longest time interval

## 24 hr upregulated
up_24h_18694499 <- df_list[["24h up"]]$Gene_Symbol

## 24 hr downregulated 
down_24h_18694499 <- df_list[["24h down"]]$Gene_Symbol 

# Checking against LUAD
print(c("Genes common to upregulated 24 hr cell line genes and LUAD_A1_TE: ", intersect(LUAD_A1_TE, up_24h_18694499)))
print(c("Genes common to upregulated 24 hr cell line genes and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, up_24h_18694499)))
print(c("Genes common to upregulated 24 hr cell line genes and LUAD_A1_TE_A2_TM: ", intersect(LUAD_A1_TE_A2_TM, up_24h_18694499)))

print(c("Genes common to downregulated 24 hr cell line genes and LUAD_A1_TE: ", intersect(LUAD_A1_TE, down_24h_18694499)))
print(c("Genes common to downregulated 24 hr cell line genes and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, down_24h_18694499)))
print(c("Genes common to downregulated 24 hr cell line genes and LUAD_A1_TE_A2_TM: ", intersect(LUAD_A1_TE_A2_TM, down_24h_18694499)))

# Checking against LUSC
print(c("Genes common to upregulated 24 hr cell line genes and LUSC_A1_TE: ", intersect(LUSC_A1_TE, up_24h_18694499)))
print(c("Genes common to upregulated 24 hr cell line genes and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, up_24h_18694499)))
print(c("Genes common to upregulated 24 hr cell line genes and LUSC_A1_TE_A2_TM: ", intersect(LUSC_A1_TE_A2_TM, up_24h_18694499)))

print(c("Genes common to downregulated 24 hr cell line genes and LUSC_A1_TE: ", intersect(LUSC_A1_TE, down_24h_18694499)))
print(c("Genes common to downregulated 24 hr cell line genes and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, down_24h_18694499)))
print(c("Genes common to downregulated 24 hr cell line genes and LUSC_A1_TE_A2_TM: ", intersect(LUSC_A1_TE_A2_TM, down_24h_18694499)))



```

## Study 2: PMID 21095227
Chronically exposed cell lines to smoke for a month
```{r}
# Read in tables
up_21095227_S2 <- read.table("../2_Gene_lists/cell_lines/21095227/21095227_S2.txt", sep = "\t", header = T)
down_21095227_S3 <- read.table("../2_Gene_lists/cell_lines/21095227/21095227_S3.txt", sep = "\t", header = T)

# Checking against LUAD
# Upregulated after 1 month exposure:
print(c("Genes common to upregulated genes in 1 month exposure to smoke and LUAD_A1_TE: ", intersect(LUAD_A1_TE, up_21095227_S2$Gene.Symbol)))
print(c("Genes common to upregulated genes in 1 month exposure to smoke and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, up_21095227_S2$Gene.Symbol)))
print(c("Genes common to upregulated genes in 1 month exposure to smoke and LUAD_A1_TE_A2_TM: ", intersect(LUAD_A1_TE_A2_TM, up_21095227_S2$Gene.Symbol)))

# Downregulated after 1 month exposure:
print(c("Genes common to downregulated genes in 1 month exposure to smoke and LUAD_A1_TE: ", intersect(LUAD_A1_TE, down_21095227_S3$Gene_Symbol)))
print(c("Genes common to downregulated genes in 1 month exposure to smoke and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, down_21095227_S3$Gene_Symbol)))
print(c("Genes common to downregulated genes in 1 month exposure to smoke and LUAD_A1_TE_A2_TM: ", intersect(LUAD_A1_TE_A2_TM, down_21095227_S3$Gene_Symbol)))

# Checking against LUSC
# Upregulated after 1 month exposure:
print(c("Genes common to upregulated genes in 1 month exposure to smoke and LUSC_A1_TE: ", intersect(LUSC_A1_TE, up_21095227_S2$Gene.Symbol)))
print(c("Genes common to upregulated genes in 1 month exposure to smoke and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, up_21095227_S2$Gene.Symbol)))
print(c("Genes common to upregulated genes in 1 month exposure to smoke and LUSC_A1_TE_A2_TM: ", intersect(LUSC_A1_TE_A2_TM, up_21095227_S2$Gene.Symbol)))

# Downregulated after 1 month exposure:
print(c("Genes common to downregulated genes in 1 month exposure to smoke and LUSC_A1_TE: ", intersect(LUSC_A1_TE, down_21095227_S3$Gene_Symbol)))
print(c("Genes common to downregulated genes in 1 month exposure to smoke and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, down_21095227_S3$Gene_Symbol)))
print(c("Genes common to downregulated genes in 1 month exposure to smoke and LUSC_A1_TE_A2_TM: ", intersect(LUSC_A1_TE_A2_TM, down_21095227_S3$Gene_Symbol)))

```

## Study 3: PMID 28898697

Cell lines were treated with cigarette smoke for 15 months, and methylation sequencing was performed
Supplementary table 1, 15 month exposure, rep 1 and rep 2 (combined), hypermethylation

```{r}
## Read in the table 
S1_28898697_15m <- read.table("../2_Gene_lists/cell_lines/28898697/28898697-s1-15months.txt", header = TRUE)

## Compare with LUAD
paste0("PMID 28898697 list and LUAD_A1_TE: ", length(intersect(LUAD_A1_TE, S1_28898697_15m$gene)))
writeLines(intersect(LUAD_A1_TE, S1_28898697_15m$gene))
paste0("PMID 28898697 list and LUAD_A1_TE_A2: ", length(intersect(LUAD_A1_TE_A2, S1_28898697_15m$gene)))

## Compare with LUSC
paste0("PMID 28898697 list and LUSC_A1_TE: ", length(intersect(LUSC_A1_TE, S1_28898697_15m$gene)))
writeLines(intersect(LUSC_A1_TE, S1_28898697_15m$gene))
paste0("PMID 28898697 list and LUSC_A1_TE_A2: ", length(intersect(LUSC_A1_TE_A2, S1_28898697_15m$gene)))

# Total number of unique genes
print("Unique genes in PMID 28898697 list:")
length(unique(S1_28898697_15m$gene))

```


## Study 4: PMID 20689807

My plan: Take the genes that are significant in NS vs SC, and not significant in NS vs SNC. These will be the genes specific to smokers with cancer. I will define it by FDR <= 0.05 Then I will check these genes against the genes of interest.

```{r}
## Read in the table
S1_20689807 <- read.table("../2_Gene_lists/irreversible_expression_airways/20689807/20689807_S1.txt", header = TRUE)

## Filter to genes with FDR <= 0.05 for SC vs NS
S1_20689807_SC_NS <- S1_20689807 %>%
  filter(fdr_SC_NS <= 0.05)

## Filter to genes with FDR <= 0.05 for SNC vs NS
S1_20689807_SNC_NS <- S1_20689807 %>%
  filter(fdr_SNC_NS <= 0.05)

## Filter to genes in SC vs NS and NOT SNC vs NS
S1_20689807_SC_NS_specific <- S1_20689807_SC_NS %>%
  filter(!gene %in% S1_20689807_SNC_NS$gene)

## Now check against the genes of interest

## Compare with LUAD
paste0("PMID 20689807 list and LUAD_A1_TE: ", length(intersect(LUAD_A1_TE, S1_20689807_SC_NS_specific$gene)))
writeLines(intersect(LUAD_A1_TE, S1_20689807_SC_NS_specific$gene))
paste0("PMID 20689807 list and LUAD_A1_TE_A2: ", length(intersect(LUAD_A1_TE_A2, S1_20689807_SC_NS_specific$gene)))
writeLines(intersect(LUAD_A1_TE_A2, S1_20689807_SC_NS_specific$gene))
paste0("PMID 20689807 list and LUAD_A1_TE_A2_TM: ", length(intersect(LUAD_A1_TE_A2_TM, S1_20689807_SC_NS_specific$gene)))
writeLines(intersect(LUAD_A1_TE_A2_TM, S1_20689807_SC_NS_specific$gene))

## Compare with LUSC
paste0("PMID 20689807 list and LUSC_A1_TE: ", length(intersect(LUSC_A1_TE, S1_20689807_SC_NS_specific$gene)))
writeLines(intersect(LUSC_A1_TE, S1_20689807_SC_NS_specific$gene))
paste0("PMID 20689807 list and LUSC_A1_TE_A2: ", length(intersect(LUSC_A1_TE_A2, S1_20689807_SC_NS_specific$gene)))
writeLines(intersect(LUSC_A1_TE_A2, S1_20689807_SC_NS_specific$gene))
paste0("PMID 20689807 list and LUSC_A1_TE_A2_TM: ", length(intersect(LUSC_A1_TE_A2_TM, S1_20689807_SC_NS_specific$gene)))
writeLines(intersect(LUSC_A1_TE_A2_TM, S1_20689807_SC_NS_specific$gene))

```


# Study 5: PMID 31015447

## Investigating whether it is feasible to compare brushings from former smokers with and without cancer
```{r}
library(dplyr)

# Read in the tables
table_1_31015447 <- read.table("../2_Gene_lists/irreversible_expression_airways/31015447_2019/31015447_table_1.txt", sep = "\t", header = T)

table_2_31015447 <- read.table("../2_Gene_lists/irreversible_expression_airways/31015447_2019/31015447_table_2.txt", sep = "\t", header = T)

# Filter table 1 to brushings from either no cancer or LUSC
table_1_31015447_LUSC_NC_brushings <- table_1_31015447 %>%
  filter(SampleType == "Brush") %>%
  filter(Previous.LUSC.History == c("LC History - LUSC","No LC History"))

# Filter table 2 to former/never smoker brushings
table_2_31015447_FS_NS_brushings <- table_2_31015447 %>%
  filter(SampleType == "Brush") %>%
  filter(Genomic_Smoking_Status == "Former/Never")

```

Just thinking about it, I don't think there is a way to deduce which of the no lung cancer samples are 'genomic' former smokers from the information here. So we might be hooped in this regard...

Could simply note that there is an upregulation of module 8 genes that they reported to be upregulated in former smokers?

The study above notes: " The Inflammatory subtype is predominated by PMLs from former smokers, but interestingly is not significantly enriched for dysplasia, despite similarly decreased expression of cilia-associated pathways, suggesting an abnormal epithelium. he Inflammatory subtype also shows increased expression of a gene module enriched for genes involved in inflammation and regulation of lymphocytes and leukocytes (Module 8). This gene module is also elevated in the Secretory subtype predominated by current smokers and increased expression of goblet cell markers. Interestingly, IL1B is part of this inflammation-related gene module, and inhibition of IL1B has recently been shown to reduce lung cancer incidence28."

Maybe I can check on the module 8 genes and see if they are among the genes that ended up in LUSC_A1_TE_A2 (therefore associated with former smoker airways).But may as well also check the other lists of interest too.

(I converted ENSEMBL IDs to gene symbols using DAVID online database)

## Checking module 8 genes in genes of interest
```{r}
library(eulerr)

# Read in genes with modules and conversion table
table_3_31015447 <- read.table("../2_Gene_lists/irreversible_expression_airways/31015447_2019/31015447_table_3.txt", sep = "\t", header = T)

table_3_31015447_gene_conversion_table <- read.table("../2_Gene_lists/irreversible_expression_airways/31015447_2019/31015477_table_3_gene_conversion.txt", sep = "\t", header = T)

# Join the ensembl IDs to the matching gene names
table_3_31015447_gene_names <- table_3_31015447 %>%
  full_join(table_3_31015447_gene_conversion_table, by = "EnsemblID")

# Select module 8
table_3_31015447_module_8 <- table_3_31015447_gene_names %>%
  filter(Gene.Module.Number == 8)

# Get the gene names (removing NA values)
module_8_genes_31015447 <- table_3_31015447_module_8$Gene[!is.na(table_3_31015447_module_8$Gene)]

# Check overlaps with the LUAD and LUSC lists

# LUAD
##LUAD_A1_TE:
print(c("Genes common to module 8 from 2019 study (PMID 31015447) and LUAD_A1_TE: ", intersect(LUAD_A1_TE, module_8_genes_31015447)))
##LUAD_A1_TE_A2:
print(c("Genes common to Barel 2023 serial sampling irreversible gene list and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, module_8_genes_31015447)))

# LUSC
##LUSC_A1_TE:
print(c("Genes common to module 8 from 2019 study (PMID 31015447) and LUSC_A1_TE: ", intersect(LUSC_A1_TE, module_8_genes_31015447)))
##LUSC_A1_TE_A2:
print(c("Genes common to Barel 2023 serial sampling irreversible gene list and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, module_8_genes_31015447)))

# Plot a euler/venn diagram for these
LUSC_A1_TE_mod8_31015447_gene_sets <- euler(list("LUSC-A1-TE" = LUSC_A1_TE, "module 8 genes (PMID 31015447)" = module_8_genes_31015447))

plot(main = "Shared genes with module 8 (PMID 31015447)",
     LUSC_A1_TE_mod8_31015447_gene_sets,
     quantities = TRUE,  # Show counts in regions
     fills = c("red", "blue", "green"),  
     alpha = 0.5, 
     edges = TRUE)

```

## Study 6: PMID 37531632
Serial sampling of small airways over 1 year of smoking cessation
```{r}

## Barel et all 2023 (Cornell lab serial sampling study)
barel_2023_irreversible_gene_list <- read.table("../2_Gene_lists/irreversible_expression_airways/37531632_barel_2023 /Table_2_truncated_37531632.txt", header = TRUE)

##LUAD_A1_TE:
print(c("Genes common to Barel 2023 serial sampling irreversible gene list and LUAD_A1_TE: ", intersect(LUAD_A1_TE, barel_2023_irreversible_gene_list$Gene)))
##LUAD_A1_TE_A2:
print(c("Genes common to Barel 2023 serial sampling irreversible gene list and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, barel_2023_irreversible_gene_list$Gene)))

##LUSC_A1_TE:
print(c("Genes common to Barel 2023 serial sampling irreversible gene list and LUSC_A1_TE: "))
writeLines(intersect(LUSC_A1_TE, barel_2023_irreversible_gene_list$Gene))
##LUSC_A1_TE_A2:
print(c("Genes common to Barel 2023 serial sampling irreversible gene list and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, barel_2023_irreversible_gene_list$Gene)))

```

# Study 7: PMID 23842454
```{r}

## Buro-Auriemma et al 2013 (smoking-related airway methylation changes)##

Buro_2013_25_hypermethylated <- read.table("../2_Gene_lists/methylation_airways/23842454/23842454_Buro_2013_25_hypermethylated.txt")

print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and LUAD_A1_TE:", intersect(LUAD_A1_TE, Buro_2013_25_hypermethylated$V1)))
## 2 hits, checking next layer
print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and LUAD_A1_TE_A2:", intersect(LUAD_A1_TE_A2, Buro_2013_25_hypermethylated$V1)))
# 1 hit, checking next layer
print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and LUAD_A1_TE_A2_TM:", intersect(LUAD_A1_TE_A2_TM, Buro_2013_25_hypermethylated$V1)))
# No hit

print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and LUSC_A1_TE: ", intersect(LUSC_A1_TE, Buro_2013_25_hypermethylated$V1)))
## 2 hits, checking next layer
print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, Buro_2013_25_hypermethylated$V1)))
# 1 hit, next layer
print(c("Genes common to Buro-Auriemma smoking-related hypermethylated gene list and LUSC_A1_TE_A2_TM: ", intersect(LUSC_A1_TE_A2_TM, Buro_2013_25_hypermethylated$V1)))
# No hits

###

Buro_2013_25_hypomethylated <- read.table("../2_Gene_lists/methylation_airways/23842454/23842454_Buro_2013_25_hypomethylated.txt")

print(c("Genes common to Buro-Auriemma 2013 smoking-related hypomethylated gene list and LUAD_A1_TE:", intersect(LUAD_A1_TE, Buro_2013_25_hypomethylated$V1)))
# 1 hit, next layer
print(c("Genes common to Buro-Auriemma 2013 smoking-related hypomethylated gene list and LUAD_A1_TE_A2:", intersect(LUAD_A1_TE_A2, Buro_2013_25_hypomethylated$V1)))
# No hits

print(c("Genes common to Buro-Auriemma 2013 smoking-related hypomethylated gene list and LUSC_A1_TE:", intersect(LUSC_A1_TE, Buro_2013_25_hypomethylated$V1)))
# 3 hits
print(c("Genes common to Buro-Auriemma 2013 smoking-related hypomethylated gene list and LUSC_A1_TE_A2:", intersect(LUSC_A1_TE_A2, Buro_2013_25_hypomethylated$V1)))
# No hits
```

###

## More studies

## Methylation in COPD airways
```{r}

## Vucic et al 2014 (link to COPD) ##

# Read in table 2 ("Differentially Methylated and Differentially Expressed Genes in Chronic Obstructive Pulmonary Disease Small Airways Previously Associated with Chronic Obstructive Pulmonary Disease")
vucic_2014_table_2 <- read.table("../2_Gene_lists/methylation_airways/24298892/24298892_vucic_2014_table_2", sep = '\t', header = TRUE)

print(c("Genes common to Vucic 2014 known COPD DMG list and LUAD_A1_TE: ", intersect(LUAD_A1_TE, vucic_2014_table_2$Symbol)))
# Nothing here, so no need to go further with LUAD. Check LUSC.
print(c("Genes common to Vucic 2014 known COPD DMG list and LUSC_A1_TE: ", intersect(LUSC_A1_TE, vucic_2014_table_2$Symbol)))
# Nothing here

###

# Read in table 3 ("Differentially Methylated and Expressed Genes Most Likely Under Epigenetic Control in Chronic Obstructive Pulmonary Disease Small Airways")
vucic_2014_table_3 <- read.table("../2_Gene_lists/methylation_airways/24298892/24298892_vucic_2014_table_3", sep = '\t', header = TRUE)

print(c("Genes common to Vucic 2014 COPD-related linked DEG-DMG list and LUAD_A1_TE: ", intersect(LUAD_A1_TE, vucic_2014_table_3$Ref_Seq)))
# Nothing

print(c("Genes common to Vucic 2014 COPD-related linked DEG-DMG  list and LUSC_A1_TE: ", intersect(LUSC_A1_TE, vucic_2014_table_3$Ref_Seq)))
# CYP4F11
# Checking the next layer
print(c("Genes common to Vucic 2014 COPD-related linked DEG-DMG  list and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, vucic_2014_table_3$Ref_Seq)))
# Not there

###


```


## Compare with study of irreversible expression changes in tumour-adjacent normal tissue
```{r}

# Read in Table 3 from Bosse et al 2012 ("Replicated probe sets upregulated by smoking and not returning to never-smoker levels within 10 years of smoking cessation in the discovery set and validated in at least one of the 2 replication sets")

bosse_2012_table_3 <- read.table("../2_Gene_lists/tumor_adjacent_normal/22659451_bosse_2012_table_3.txt", header = TRUE)


##LUAD_A1_TE
print(c("Genes common to Bosse 2012 list of genes persistently upregulated in tumor-adjacent normal tissue 10 years after smoking cessation, and the LUAD-A1-TE list:", intersect(LUAD_A1_TE, bosse_2012_table_3$Gene_symbol)))
##LUAD_A1_TE_A2
print(c("Genes common to Bosse 2012 list of genes persistently upregulated in tumor-adjacent normal tissue 10 years after smoking cessation, and the LUAD-A1-TE-A2 list:", intersect(LUAD_A1_TE_A2, bosse_2012_table_3$Gene_symbol)))


##LUSC_A1_TE
print(c("Genes common to Bosse 2012 list of genes persistently upregulated in tumor-adjacent normal tissue 10 years after smoking cessation, and LUSC_A1_TE:", intersect(LUSC_A1_TE, bosse_2012_table_3$Gene_symbol)))
##LUSC_A1_TE_A2
print(c("Genes common to Bosse 2012 list of genes persistently upregulated in tumor-adjacent normal tissue 10 years after smoking cessation, and LUSC_A1_TE_A2:", intersect(LUSC_A1_TE_A2, bosse_2012_table_3$Gene_symbol)))

```


## Compare linked gene lists with studies related to irreversible expression changes in airways
```{r}

## Chari et al 2007 (Lam lab SAGE study) ##

raj_2007_irreversible_gene_list <- read.table("../2_Gene_lists/irreversible_expression_airways/17727719_raj_2007/17727719_raj_2007_irreversible_gene_list.txt")
colnames(raj_2007_irreversible_gene_list) <- "Gene_Symbol"


##LUAD_A1_TE
print(c("Genes common to Chari 2007 irreversible gene list and LUAD_A1_TE: ", intersect(LUAD_A1_TE, raj_2007_irreversible_gene_list$Gene)))
##LUAD_A1_TE_A2
print(c("Genes common to Chari 2007 irreversible gene list and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, raj_2007_irreversible_gene_list$Gene)))

##LUSC_A1_TE
print(c("Genes common to Chari 2007 irreversible gene list and LUSC_A1_TE: ", intersect(LUSC_A1_TE, raj_2007_irreversible_gene_list$Gene)))
##LUSC_A1_TE_A2
print(c("Genes common to Chari 2007 irreversible gene list and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, raj_2007_irreversible_gene_list$Gene)))

## (For sake of completeness do the reference study later. Also make venn diagrams of this.)
```

Maybe it would make a lot more sense to look at these module 8 genes in TCGA-LUSC of former smokers? Anyway, there you are. Quite a few overlap with LUSC_A1_TE but none with LUSC_A1_TE_A2.

STRING shows that this list is really enriched for immune related stuff:
https://version-12-0.string-db.org/cgi/network?networkId=bLP09oa886uo

Note: I haven't been paying attention to sign for the genes of interest here. I should really be checking on that, it could indicate a reversal event as well since some are up and then down later...

# Vanessa's darkolivegreen module (associated with former smoker airways!(?))
```{r}
vanessa_darkolivegreen_module_20250602 <- readLines("../2_Gene_lists/vanessa_darkolivegreen_module_20250602.txt")

# LUAD
print(c("Genes common to Vanessa's darkolivegreen module and LUAD_A1_TE: ", intersect(LUAD_A1_TE, vanessa_darkolivegreen_module_20250602)))
print(c("Genes common to Vanessa's darkolivegreen module and LUAD_A1_TE_A2: ", intersect(LUAD_A1_TE_A2, vanessa_darkolivegreen_module_20250602)))

# LUSC
print(c("Genes common to Vanessa's darkolivegreen module and LUSC_A1_TE: ", intersect(LUSC_A1_TE, vanessa_darkolivegreen_module_20250602)))
print(c("Genes common to Vanessa's darkolivegreen module and LUSC_A1_TE_A2: ", intersect(LUSC_A1_TE_A2, vanessa_darkolivegreen_module_20250602)))

```

STRING link for the LUSC ones:
https://version-12-0.string-db.org/cgi/network?networkId=b8Kfbonk01Wz

This might be kinda self-selecting though since she got these partly by investigating TCGA. And if they don't carry on to the FS list then it might have limited info? Interesting anyway.


