---
title: "Tumor_stage_correlation"
output: html_document
date: "2025-06-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


I want to check the genes of interest for tumor stage correlation to particularly see if any are associated with early stage tumors.

## Load libraries
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
```


## Read in genes of interest
```{r}
LUAD_A1_TE <- read.table("../../2_Outputs/4_Linked_genes/LUAD_A1_TE_20250514.txt")
LUAD_A1_TE_A2<- read.table("../../2_Outputs/4_Linked_genes/LUAD_A1_TE_A2_20250514.txt")
LUAD_A1_TE_A2_TM <- read.table("../../2_Outputs/4_Linked_genes/LUAD_A1_TE_A2_TM_20250610.txt")

LUSC_A1_TE <- read.table("../../2_Outputs/4_Linked_genes/LUSC_A1_TE_20250514.txt")
LUSC_A1_TE_A2 <- read.table("../../2_Outputs/4_Linked_genes/LUSC_A1_TE_A2_20250514.txt")
LUSC_A1_TE_A2_TM <- read.table("../../2_Outputs/4_Linked_genes/LUSC_A1_TE_A2_TM_20250610.txt")
```

# LUAD

## Download clinical data
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)

query_clinical <- GDCquery(
  project = "TCGA-LUAD",
  data.category = "Clinical",
  data.type = "Clinical Supplement"
)

# Check the available files
query_clinical$results[[1]] %>% head()

## I see that some are not BCR XML files, so I will try to remove these
query_clinical$results[[1]] <- query_clinical$results[[1]] %>%
  filter(data_format == "BCR XML")

## Download data
GDCdownload(query_clinical, directory = "../../GDC_data")
clinical_data <- GDCprepare_clinic(query_clinical, clinical.info = "patient", directory = "../../GDC_data")

head(clinical_data)
```

### Download expression data
```{r}
query <- GDCquery(project = "TCGA-LUAD",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  sample.type = "Primary Tumor",
                  workflow.type = "STAR - Counts")

GDCdownload(query, directory = "../../GDC_data")

data <- GDCprepare(query, directory = "../../GDC_data")

counts <- as.data.frame(assay(data))  # Extracting the count matrix (these are supposedly raw counts)
head(counts)  # Viewing the first few rows (genes) and columns (samples)

gene_info <- as.data.frame(rowData(data))
head(gene_info)  # Preview the first few genes and their annotations

sample_info <- as.data.frame(colData(data))
head(sample_info)  # Preview sample metadata

```

## Match clinical to expression data
```{r}
library(edgeR)

# Filter out lowly expressed genes
keep <- filterByExpr(counts)
counts_filtered <- counts[keep, ]

# Make sure sample barcodes match clinical format
colnames(counts_filtered) <- substr(colnames(counts_filtered), 1, 12) ## Confirm what this does

# Keep only matching samples in both data sets
matching_samples <- intersect(colnames(counts_filtered), clinical_data$bcr_patient_barcode)
counts_filtered <- counts_filtered[, matching_samples]
clinical_data <- clinical_data[clinical_data$bcr_patient_barcode %in% matching_samples, ]

# Order both the same
clinical_data <- clinical_data[match(colnames(counts_filtered), clinical_data$bcr_patient_barcode), ]
stopifnot(all(colnames(counts_filtered) == clinical_data$bcr_patient_barcode))

# Seems there are still 8 extra samples in the counts data, so remove those
samples_to_remove <- setdiff(colnames(counts_filtered), clinical_data$bcr_patient_barcode)
counts_filtered <- counts_filtered %>%
  select(-all_of(samples_to_remove))

```

## Normalize raw counts (logCPM for correlation)
```{r}
dge <- DGEList(counts = counts_filtered)
dge <- calcNormFactors(dge)
logCPM <- cpm(dge, log = TRUE)

## Store this LUAD specific logCPM
LUAD_logCPM <- logCPM
```

## Clean and parse tumor stage
```{r}
stage <- clinical_data$stage_event_pathologic_stage
stage_clean <- gsub("stage ", "", tolower(stage))
stage_clean <- factor(stage_clean, 
                      levels = c("i", "ia", "ib", "ii", "iia", "iib", 
                                 "iii", "iiia", "iiib", "iiic", "iv"),
                      ordered = TRUE)

# Drop samples with missing stage
valid_idx <- which(!is.na(stage_clean))
logCPM <- logCPM[, valid_idx]
stage_clean <- stage_clean[valid_idx]
clinical_data <- clinical_data[valid_idx, ]

# Store this LUAD specific stage info and clinical data
LUAD_stage_clean <- stage_clean
LUAD_clinical_data <- clinical_data
```

## Correlate gene expression with tumor stage
```{r}
logCPM <- LUAD_logCPM
stage_clean <- LUAD_stage_clean
clinical_data <- LUAD_clinical_data

genes_of_interest <- LUAD_A1_TE$Gene

results <- data.frame(Gene = character(), SpearmanRho = numeric(), p_value = numeric())

for (gene in genes_of_interest) {
  gene_id <- rownames(gene_info)[gene_info$gene_name == gene]
  if (length(gene_id) == 0 || !(gene_id %in% rownames(logCPM))) next
  
  expr <- as.numeric(logCPM[gene_id, ])
  cor_test <- cor.test(expr, as.numeric(stage_clean), method = "spearman")
  
  results <- rbind(results, data.frame(
    Gene = gene,
    SpearmanRho = cor_test$estimate,
    p_value = cor_test$p.value
  ))
}

# Adjust p-values
results$FDR <- p.adjust(results$p_value, method = "BH")

## Give better label 
LUAD_A1_TE_tumor_stage_spearman <- results

## Add column for log2FC in the original A1_TE list
LUAD_A1_TE_tumor_stage_spearman <- LUAD_A1_TE_tumor_stage_spearman %>%
  left_join(., LUAD_A1_TE[,c("Gene", "log2FC_A1")], by = "Gene")

# save (unfiltered) results

#write.table(LUAD_A1_TE_tumor_stage_spearman, "../../2_Outputs/LUAD_A1_TE_tumor_stage_spearman_20250613.txt")

```

## Filtering and checking results
```{r}

## Filter results for FDR <= 0.05
LUAD_A1_TE_tumor_stage_spearman_filtered <- LUAD_A1_TE_tumor_stage_spearman[LUAD_A1_TE_tumor_stage_spearman$FDR<=0.05,]


## Check which of these genes were in LUAD_A1_TE_A2
LUAD_A1_TE_A2_tumor_stage_spearman_filtered <- LUAD_A1_TE_tumor_stage_spearman_filtered %>%
  filter(Gene %in% LUAD_A1_TE_A2$Gene)

## Check which of these genes were in LUAD_A1_TE_A2_TM
LUAD_A1_TE_A2_TM_tumor_stage_spearman_filtered <- LUAD_A1_TE_tumor_stage_spearman_filtered %>%
  filter(Gene %in% LUAD_A1_TE_A2_TM$Gene)


# Now I am wondering specifically about genes that are be more highly expressed in early tumors (so negative rho) but with higher expression in smoker airways and tumor vs normal (so positive log2FC_A1). I would postulate that maybe these genes could be important in the carcinogenesis process but not in the later tumor progression. Of course there are other possibilities and genes with other signs could be important too.
LUAD_A1_TE_tumor_stage_spearman_filtered_down_up <- LUAD_A1_TE_tumor_stage_spearman_filtered %>%
  filter(sign(SpearmanRho) == -1) %>%
  filter(sign(log2FC_A1) == 1)
# These don't seem to be in any common pathways, so maybe a red herring, idk.

# I think you could tell similar stories about genes with any combination of those signs, so I will check the other combinations.
LUAD_A1_TE_tumor_stage_spearman_filtered_up_down <- LUAD_A1_TE_tumor_stage_spearman_filtered %>%
  filter(sign(SpearmanRho) == 1) %>%
  filter(sign(log2FC_A1) == -1)

LUAD_A1_TE_tumor_stage_spearman_filtered_up_up <- LUAD_A1_TE_tumor_stage_spearman_filtered %>%
  filter(sign(SpearmanRho) == 1) %>%
  filter(sign(log2FC_A1) == 1)

LUAD_A1_TE_tumor_stage_spearman_filtered_down_down <- LUAD_A1_TE_tumor_stage_spearman_filtered %>%
  filter(sign(SpearmanRho) == -1) %>%
  filter(sign(log2FC_A1) == -1)

```


Now I want to try plotting some of the genes that are really significant and have negative correlation. Those are the ones we would say might be associated with early cancer / carcinogenesis, which is why we would see them in airways of people who do not yet have cancer. (I also try plotting some other guys.)

## Plot expression by tumor stage
```{r}

library(ggplot2)
library(dplyr)
library(tidyr)

logCPM <- LUAD_logCPM
stage_clean <- LUAD_stage_clean
clinical_data <- LUAD_clinical_data

# Pick one or more genes of interest to plot
#genes_to_plot <- c("FCRL2","TTC16","CLNK") # Highly negatively correlated ones
#genes_to_plot <- c("GMDS", "NQO1") ## In TM and both positive rho/positive A1 log2FC
genes_to_plot <- "SLC2A1" # The one with lowest p value by far

plot_data <- data.frame()
label_data <- data.frame()

for (gene in genes_to_plot) {
  gene_id <- rownames(gene_info)[gene_info$gene_name == gene]
  if (length(gene_id) == 0 || !(gene_id %in% rownames(logCPM))) next
  
  df_plot <- data.frame(
    Expression = as.numeric(logCPM[gene_id, ]),
    Stage = stage_clean,
    Gene = gene
  )
  
  df_labels <- LUAD_A1_TE_tumor_stage_spearman[LUAD_A1_TE_tumor_stage_spearman$Gene ==gene, ]
  
  plot_data <- rbind(plot_data, df_plot)
  label_data <- rbind(label_data, df_labels)
}

# Add numeric stage to plot_data
plot_data$Stage_num <- as.numeric(plot_data$Stage)

## Plot boxplots with trendline
ggplot(plot_data, aes(x = Stage, y = Expression)) +
  geom_boxplot(outlier.shape = NA, fill = "lightblue", color = "black", width = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  geom_smooth(aes(x = Stage_num, y = Expression), method = "lm", color = "black", se = TRUE, inherit.aes = FALSE) +
  facet_wrap(~ Gene, scales = "free_y") +
  
  geom_text(
    data = label_data,
    aes(label = paste0("rho = ", round(SpearmanRho, 2),
                       "\np = ", signif(p_value, 2),
                       "\nFDR = ", signif(FDR, 2))),
    x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, inherit.aes = FALSE,
    size = 3.5
  ) +
  
  labs(title = "Expression by Tumor Stage with Trendline",
       x = "Tumor Stage", y = "logCPM Expression") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



## Violin plot version just for a try


ggplot(plot_data, aes(x = Stage, y = Expression)) +
  geom_violin(fill = "lightblue", color = "black", width = 0.9, scale = "width") +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  geom_smooth(aes(x = Stage_num, y = Expression), method = "lm", color = "black", se = TRUE, inherit.aes = FALSE) +
  facet_wrap(~ Gene, scales = "free_y") +

  geom_text(
    data = label_data,
    aes(label = paste0("rho = ", round(SpearmanRho, 2),
                       "\np = ", signif(p_value, 2),
                       "\nFDR = ", signif(FDR, 2))),
    x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, inherit.aes = FALSE,
    size = 3.5
  ) +

  labs(title = "Expression by Tumor Stage (Violin) with Trendline",
       x = "Tumor Stage", y = "logCPM Expression") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```







####################





## LUSC

## Download clinical data
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)

query_clinical <- GDCquery(
  project = "TCGA-LUSC",
  data.category = "Clinical",
  data.type = "Clinical Supplement"
)

# Check the available files
query_clinical$results[[1]] %>% head()

## I see that some are not BCR XML files, so I will try to remove these
query_clinical$results[[1]] <- query_clinical$results[[1]] %>%
  filter(data_format == "BCR XML")

## Download data
GDCdownload(query_clinical, directory = "../../GDC_data")
clinical_data <- GDCprepare_clinic(query_clinical, clinical.info = "patient", directory = "../../GDC_data")

head(clinical_data)
```

### Download expression data
```{r}
query <- GDCquery(project = "TCGA-LUSC",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification",
                  sample.type = "Primary Tumor",
                  workflow.type = "STAR - Counts")

GDCdownload(query, directory = "../../GDC_data")

data <- GDCprepare(query, directory = "../../GDC_data")

counts <- as.data.frame(assay(data))  # Extracting the count matrix (these are supposedly raw counts)
head(counts)  # Viewing the first few rows (genes) and columns (samples)

gene_info <- as.data.frame(rowData(data))
head(gene_info)  # Preview the first few genes and their annotations

sample_info <- as.data.frame(colData(data))
head(sample_info)  # Preview sample metadata

```

## Match clinical to expression data
```{r}
library(edgeR)

# Filter out lowly expressed genes
keep <- filterByExpr(counts)
counts_filtered <- counts[keep, ]

# Make sure sample barcodes match clinical format
colnames(counts_filtered) <- substr(colnames(counts_filtered), 1, 12) ## Confirm what this does

# Keep only matching samples in both data sets
matching_samples <- intersect(colnames(counts_filtered), clinical_data$bcr_patient_barcode)
counts_filtered <- counts_filtered[, matching_samples]
clinical_data <- clinical_data[clinical_data$bcr_patient_barcode %in% matching_samples, ]

# Order both the same
clinical_data <- clinical_data[match(colnames(counts_filtered), clinical_data$bcr_patient_barcode), ]
stopifnot(all(colnames(counts_filtered) == clinical_data$bcr_patient_barcode))

# Seems there are still 8 extra samples in the counts data, so remove those
samples_to_remove <- setdiff(colnames(counts_filtered), clinical_data$bcr_patient_barcode)
counts_filtered <- counts_filtered %>%
  select(-all_of(samples_to_remove))

```

## Normalize raw counts (logCPM for correlation)
```{r}
dge <- DGEList(counts = counts_filtered)
dge <- calcNormFactors(dge)
logCPM <- cpm(dge, log = TRUE)

# Store as lusc specific
LUSC_logCPM <- logCPM
```

## Clean and parse tumor stage
```{r}
stage <- clinical_data$stage_event_pathologic_stage
stage_clean <- gsub("stage ", "", tolower(stage))
stage_clean <- factor(stage_clean, 
                      levels = c("i", "ia", "ib", "ii", "iia", "iib", 
                                 "iii", "iiia", "iiib", "iiic", "iv"),
                      ordered = TRUE)

# Drop samples with missing stage
valid_idx <- which(!is.na(stage_clean))
logCPM <- logCPM[, valid_idx]
stage_clean <- stage_clean[valid_idx]
clinical_data <- clinical_data[valid_idx, ]

# store as lusc specific
LUSC_stage_clean <- stage_clean
LUSC_clinical_data <- clinical_data
```

## Correlate gene expression with tumor stage
```{r}
stage_clean <- LUSC_stage_clean
clinical_data <- LUSC_clinical_data
logCPM <- LUSC_logCPM

genes_of_interest <- LUSC_A1_TE$Gene

results <- data.frame(Gene = character(), SpearmanRho = numeric(), p_value = numeric())

for (gene in genes_of_interest) {
  gene_id <- rownames(gene_info)[gene_info$gene_name == gene]
  if (length(gene_id) == 0 || !(gene_id %in% rownames(logCPM))) next
  
  expr <- as.numeric(logCPM[gene_id, ])
  cor_test <- cor.test(expr, as.numeric(stage_clean), method = "spearman")
  
  results <- rbind(results, data.frame(
    Gene = gene,
    SpearmanRho = cor_test$estimate,
    p_value = cor_test$p.value
  ))
}

## Add column for log2FC in the original A1_TE list
LUSC_A1_TE_tumor_stage_spearman <- LUSC_A1_TE_tumor_stage_spearman %>%
  left_join(., LUSC_A1_TE[,c("Gene", "log2FC_A1")], by = "Gene")

# save (unfiltered) results
#write.table(LUSC_A1_TE_tumor_stage_spearman, "../../2_Outputs/LUSC_A1_TE_tumor_stage_spearman_20250613.txt")

```

## Filtering and checking results
```{r}

## Filter results for FDR <= 0.05
LUSC_A1_TE_tumor_stage_spearman_filtered <- LUSC_A1_TE_tumor_stage_spearman[LUSC_A1_TE_tumor_stage_spearman$FDR<=0.05,]


## Check which of these genes were in LUSC_A1_TE_A2
LUSC_A1_TE_A2_tumor_stage_spearman_filtered <- LUSC_A1_TE_tumor_stage_spearman_filtered %>%
  filter(Gene %in% LUSC_A1_TE_A2$Gene)

## None of the other checks I did for LUAD are worth doing because it is such a small list 

## Actually, maybe of note: MTHFD2 and PYCR1 are the only genes with positive rho for tumor stage and positive log2FC in LUSC_A1_TE,
## and these have a common function: https://version-12-0.string-db.org/cgi/network?networkId=b99p5bXvGXGo

## And a really interesting paper that mentions them in conjunction related to tumorgenesis:
## PMID: 24451681
## Also maybe neat: PMID: 34121323

```


Now I want to try plotting some of the genes that are of interest

## Plot expression by tumor stage
```{r}

stage_clean <- LUSC_stage_clean
clinical_data <- LUSC_clinical_data
logCPM <- LUSC_logCPM


library(ggplot2)
library(dplyr)
library(tidyr)

# Pick one or more genes of interest to plot
genes_to_plot <- c("MTHFD2", "PYCR1")

plot_data <- data.frame()
label_data <- data.frame()

for (gene in genes_to_plot) {
  gene_id <- rownames(gene_info)[gene_info$gene_name == gene]
  if (length(gene_id) == 0 || !(gene_id %in% rownames(logCPM))) next
  
  df_plot <- data.frame(
    Expression = as.numeric(logCPM[gene_id, ]),
    Stage = stage_clean,
    Gene = gene
  )
  
  df_labels <- LUSC_A1_TE_tumor_stage_spearman[LUSC_A1_TE_tumor_stage_spearman$Gene ==gene, ]
  
  plot_data <- rbind(plot_data, df_plot)
  label_data <- rbind(label_data, df_labels)
}

# Add numeric stage to plot_data
plot_data$Stage_num <- as.numeric(plot_data$Stage)

## Plot boxplots with trendline
ggplot(plot_data, aes(x = Stage, y = Expression)) +
  geom_boxplot(outlier.shape = NA, fill = "lightblue", color = "black", width = 0.6) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  geom_smooth(aes(x = Stage_num, y = Expression), method = "lm", color = "black", se = TRUE, inherit.aes = FALSE) +
  facet_wrap(~ Gene, scales = "free_y") +
  
  geom_text(
    data = label_data,
    aes(label = paste0("rho = ", round(SpearmanRho, 2),
                       "\np = ", signif(p_value, 2),
                       "\nFDR = ", signif(FDR, 2))),
    x = Inf, y = Inf, hjust = 1.1, vjust = 1.1, inherit.aes = FALSE,
    size = 3.5
  ) +
  
  labs(title = "Expression by Tumor Stage with Trendline",
       x = "Tumor Stage", y = "logCPM Expression") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

It seems strange that there are so many fewer genes here than with LUAD. I checked the FDR math and it's working properly. I guess that's just the way it is.





